What to simulate? Inferring the right direction for mental rotation

### 2014-05-01 man/cogsci-2014/cogsci-2014.tex
Last few small changes

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. We found that three models based on previous
  hypotheses in the literature were unable to account for several
  aspects of human behavior. The fourth is based on the idea
  \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu}, which is a
  strategy of choosing actions that will provide the most
  informative. This last model provides a plausible account of how
  people use mental rotation, where the other models do not. Based on
  these results, we suggest that the question of ``what to simulate?''
  is more difficult than has previously been assumed, and that an
  active learning approach holds promise for uncovering the answer.

@@

If people are not computing the rotation beforehand, what might they
be doing? To answer this question, we perform a rational analysis of
the problem of mental rotation
\cite{Marr:1983to,anderson90,Shepard:1987tt}. At the computational
level, we can say that the \textit{problem} is to determine which
spatial transformations an object has undergone based on two images of
that object (which do not include information about point
correspondences). At the algorithmic level, we are constrained by the
notion that mental images must be transformed in an analog manner (or
in a way that is approximately analog), and that mental images are
time-consuming and effortful to generate. Thus, the \textit{goal} is
to make this determination while performing a minimum amount of
computation (i.e., as few rotations as possible).

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$: $p(\Xb\ \vert\ \Xa, \theta, h):=
S(\Xb, f(\Xa, h, \theta))$.  We do not know which vertices of $\Xb$
correspond to which vertices of $\Xa$, so the similarity $S$ must
marginalize over the set of possible mappings. For brevity, let
$\mathbf{X}_m=\M\cdot{}f(\Xa, h, \theta)$ where $\M$ is a permutation
matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{2N} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $2N$ is the total number of possible mappings,\footnote{It is
  $2N$ and not $N^2$ because, in polar coordinates, vertices are
  always connected to their two nearest neighbors in the $\theta$
  dimension.}and $\sigma_S^2=0.15$ is the variance of the
similarity. Example similarity curves are shown in
Figure~\ref{fig:shapes}.

@@

The average RT across all correctly-judged stimuli was \ExpTime{} ms;
the full histogram of RTs can be seen in
Figure~\ref{fig:histograms}. The minimum angle of rotation was
significantly rank-order (Spearman) correlated with average
per-stimulus RTs, both for ``flipped'' (\ExpThetaTimeCorrFlipped{})
and ``same'' pairs (\ExpThetaTimeCorrSame{}). While this replicates
the general result of previous experiments
\cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are not as linear
(Figure~\ref{fig:response-time-accuracy}).\\
\indent The average accuracy across all stimuli was \ExpAccuracy{},
though there were \ExpNumChance{} stimuli (out of 720) for which
people were not above chance.  The minimum angle was also correlated
with participants' average per-stimulus accuracy, though much more so
for ``same'' pairs (\ExpThetaAccuracyCorrSame{}) than ``flipped''
pairs (\ExpThetaAccuracyCorrFlipped{}). This is the same
result found both by \citeA{Cooper:1975wp} and \citeA{Gardony:2013gn}.\\
\indent There was a significant effect of trial number both on RT
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These effects may
have contributed to the not-quite-linearity of the human RTs; future
work should collect more data per participant.

@@

\bibliographystyle{apacite}
\renewcommand{\bibliographytypesize}{\small}
\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
\bibliography{references}

>>>>

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. We found that three models based on previous
  hypotheses in the literature were unable to account for several
  aspects of human behavior. The fourth is based on the idea
  \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu}, which is a
  strategy of choosing actions that will provide the most
  information. This last model provides a plausible account of how
  people use mental rotation, where the other models do not. Based on
  these results, we suggest that the question of ``what to simulate?''
  is more difficult than has previously been assumed, and that an
  active learning approach holds promise for uncovering the answer.

@@

If people are not computing the rotation beforehand, what might they
be doing? To answer this question, we perform a rational analysis of
the problem of mental rotation
\cite{Marr:1983to,Anderson:1990,Shepard:1987tt}. At the computational
level, we can say that the \textit{problem} is to determine which
spatial transformations an object has undergone based on two images of
that object (which do not include information about point
correspondences). At the algorithmic level, we are constrained by the
notion that mental images must be transformed in an analog manner (or
in a way that is approximately analog), and that mental images are
time-consuming and effortful to generate. Thus, the \textit{goal} is
to make this determination while performing a minimum amount of
computation (i.e., as few rotations as possible).

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$: $p(\Xb\ \vert\ \Xa, \theta, h):=
S(\Xb, f(\Xa, h, \theta))$.  We do not know which vertices of $\Xb$
correspond to which vertices of $\Xa$, so the similarity $S$ must
marginalize over the set of possible mappings. For brevity, let
$\mathbf{X}_m=\M\cdot{}f(\Xa, h, \theta)$ where $\M$ is a permutation
matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{2N} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $2N$ is the total number of possible mappings,\footnote{It is
  $2N$ and not $N^2$ because, in polar coordinates, vertices are
  always connected to their two nearest neighbors in the $\theta$
  dimension.} and $\sigma_S^2=0.15$ is the variance of the
similarity. Example similarity curves are shown in
Figure~\ref{fig:shapes}.

@@

The average RT across all correctly-judged stimuli was \ExpTime{}; the
full histogram of RTs can be seen in Figure~\ref{fig:histograms}. The
minimum angle of rotation was significantly rank-order (Spearman)
correlated with average per-stimulus RTs, both for ``flipped''
(\ExpThetaTimeCorrFlipped{}) and ``same'' pairs
(\ExpThetaTimeCorrSame{}). While this replicates the general result of
previous experiments \cite<e.g.,>{Shepard1971,Cooper:1975wp}, our
results are not as linear
(Figure~\ref{fig:response-time-accuracy}).\\
\indent The average accuracy across all stimuli was \ExpAccuracy{},
though there were \ExpNumChance{} stimuli (out of 720) for which
people were not above chance.  The minimum angle was also correlated
with participants' average per-stimulus accuracy, though much more so
for ``same'' pairs (\ExpThetaAccuracyCorrSame{}) than ``flipped''
pairs (\ExpThetaAccuracyCorrFlipped{}). This is the same
result found both by \citeA{Cooper:1975wp} and \citeA{Gardony:2013gn}.\\
\indent There was a significant effect of trial number both on RT
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These effects may
have contributed to the not-quite-linearity of the human RTs; future
work should collect more data per participant.

@@

\bibliographystyle{apacite}
\renewcommand{\bibliographytypesize}{\small}
\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
\vspace{-0.27cm}
\bibliography{references}
